---
title: "Lab 6 | Advanced Visualization Tools & Customization"
subtitle: "ST 437 Data Visualization"
author: "Student Name"
date: today
date-format: long
format: html
execute: 
  echo: true
  warning: false
---

## Learning Objectives

Pending

### Overview

Spatial visualization is crucial for analyzing geographic data, allowing for the representation of information on maps. This lab will guide you through creating static, interactive, and publication-ready maps using R.

## Example 1 | Static Map with `ggplot2` and `sf`

### Introduction

Static maps are powerful tools for presenting spatial data in a clear and concise manner. In this example, we’ll visualize global population data using the World Bank’s World Development Indicators dataset.

### Step 1 | Load and Explore Spatial Data

We’ll start by loading the World Development Indicators dataset, focusing on population data, and then join it with a spatial dataset containing country boundaries.

```{r}
library(sf)
library(ggplot2)
library(dplyr)
library(readr)
```
```{r}
#| eval: false
# Load necessary libraries
library(sf)
library(dplyr)
library(readr)
library(rnaturalearth)  # Package for global country boundaries

# Load world boundaries shapefile (global country boundaries)
world <- ne_countries(scale = "medium", returnclass = "sf")

# Specify the data URL using HTTPS
url <- "https://raw.githubusercontent.com/datasets/population/master/data/population.csv"

cat("Downloading the data ...\n")
download.file(url, "population.csv")

# Read the population data into R
pop_data <- read_csv("population.csv", show_col_types = FALSE)

# Filter for the most recent year available
pop_recent <- pop_data %>%
  filter(Year == max(Year))

# Check the structure of both datasets
glimpse(pop_recent)
glimpse(world)

# Join the population data with the world boundaries using country codes
world_pop <- world %>%
  left_join(pop_recent, by = c("iso_a3" = "Country Code"))

# Save the cleaned and joined dataset
#saveRDS(world_pop, "world_population_data.rds")

cat("Dataset saved as 'world_population_data.rds'.\n")

# Load the saved RDS file
world_pop <- readRDS("world_population_data.rds")

# Display the structure of the dataset
str(world_pop)

# Preview the first few rows of the dataset
head(world_pop)

# Show a summary of the dataset
summary(world_pop)

glimpse(world_pop)
```

```{webr-r}
# Create a static map
ggplot(data = world_pop) +
  geom_sf(aes(fill = Population), color = "black") +
  scale_fill_viridis_c(option = "plasma", na.value = "lightgray") +
  labs(
    title = "Global Population Distribution",
    fill = "Population",
    caption = "Data Source: World Bank and Natural Earth"
  ) +
  theme_minimal()

```

### Step 2 | Basic Map Plotting

Let’s create a static map showing the population distribution across the world.

```{r}
# Create a basic population map
ggplot(data = world_pop) +
  geom_sf(aes(fill = Population), color = "black") +
  labs(
    title = "Global Population Distribution",
    fill = "Population",
    caption = "Data Source: World Bank and Natural Earth"
  ) +
  theme_minimal()
```

### Step 3 | Adding Layers and Customization

We can further enhance the map by customizing the color scale, adding labels, and modifying the theme.

```{r}
# Enhance the population map with customizations
ggplot(data = world_pop) +
  geom_sf(aes(fill = Population), color = "black") +
  scale_fill_viridis_c(option = "plasma", na.value = "lightgray") +
  labs(
    title = "Enhanced Global Population Distribution",
    fill = "Population",
    caption = "Data Source: World Bank and Natural Earth"
  ) +
  theme(
    panel.background = element_rect(fill = "lightblue"),
    panel.grid.major = element_line(color = "white"),
    text = element_text(family = "Georgia"),
    plot.title = element_text(size = 14, face = "bold")
  )
```

## Example 2: Interactive Map with `leaflet`

### Introduction

Interactive maps allow users to explore spatial data dynamically, making them valuable for presentations and web applications. We’ll use the `leaflet` package to create an interactive map.

### Step 1: Basic Interactive Map

We’ll create an interactive map with markers for specific locations.

```{r}
# Load the leaflet library
library(leaflet)

# Create a basic interactive map
m <- leaflet() %>%
  addTiles() %>%  # Add default OpenStreetMap tiles
  addMarkers(lng = c(-122.4194, -74.0060), lat = c(37.7749, 40.7128),
             popup = c("San Francisco", "New York"))

# Print the map
m
```

### Step 2: Adding Multiple Layers and Features

Let’s add more layers and customize the map further.

```{r}
# Enhanced interactive map with additional layers
m <- leaflet() %>%
  addTiles() %>%  # Base layer
  addProviderTiles(providers$Stamen.Toner) %>%  # Additional layer
  addMarkers(lng = c(-122.4194, -74.0060), lat = c(37.7749, 40.7128),
             popup = c("San Francisco", "New York")) %>%
  addCircles(lng = -122.4194, lat = 37.7749, radius = 50000, color = "blue", fillOpacity = 0.4) %>%
  addPolygons(lng = c(-74.1, -74.0, -73.9), lat = c(40.7, 40.8, 40.7),
              color = "red", fillOpacity = 0.3)

# Print the enhanced map
m
```

### Summary

Interactive maps with `leaflet` are ideal for web-based presentations, providing users with dynamic and engaging ways to explore spatial data.


